---
title: PalaeOpen Training School
subtitle: Setup
format:
  revealjs:
    theme: simple
    code-line-numbers: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: Gavin Simpson
    orcid: 0000-0002-9084-8413
    email: gavin@anivet.au.dk
    affiliations: Aarhus University
date: last-modified
title-slide-attributes:
  data-background-color: "#003d73"
presenter:
  name: Gavin Simpson
  institute: Department of Animal & Veterinary Sciences
knitr:
  opts_chunk: 
    echo: true
    message: false
    fig.align: center
    fig.height: 6
    fig.width: 10.66666
    collapse: true
highlight-style: github-dark
---

## GitHub Repo

Open the GitHub repo in your browser

### github.com/gavinsimpson/palaeopen-sofia-2026

## Create a folder

Using your file manager

* create a folder to store the teaching materials

* ideally create it somewhere familiar

* ideally no spaces in the folder name or path to it

This folder will be added to each day

## Install *usethis* 📦

```{r}
#| label: install-usethis
#| eval: false
install.packages("usethis")
```

## Open RStudio

* Start RStudio

* Go to Session menu

* Select **Select working directory** > **Choose Directory...**

* Use the file browser to navigate to the folder you created earlier

* Make sure you are in that folder, then select Open

## Download materials

Run the code below to download the materials

```{r}
#| label: usethis-use-course
#| eval: false
usethis::use_course(
  "gavinsimpson/palaeopen-sofia-2026",
  destdir = "./"
)
```

Answer the questions; it is OK to delete the downloaded Zip when asked.

## RStudio restarts

Once everything is completed, **usethis** will automatically restart RStudio in the folder containing the downloaded materials
